---
title: "Extra Slides"
format: html
---

## Binomial Example 

$$y_i \sim binomial(n, p_i) $$

$$ \text{logit} (p_i) = \beta_0 + \beta_1x_i $$
$$r_i = \Phi^{-1}(u_i)$$

$$ u_i \sim Uniform(a_i, b_j] $$

$$a_i = lim_{y \uparrow y_i} F(y; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k-1 \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$


$$b_i = F(y_i; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$


## Consider Groups

This is the 'wrong model' (missing 'environment'), but the plot looks okay

```{r, echo=TRUE}
wrongModel2 <- glmer(observedResponse ~ 1 + (1|group), 
                     family = "poisson", data = testData)
res2 <- simulateResiduals(wrongModel2, plot = TRUE)
```

## Consider Groups

When plotted by the missing group:

```{r, echo=TRUE}
plotResiduals(res2, form = testData$Environment1)
```

## Standard Residual Plots

```{r}
data(warpbreaks)
m1 <- lm(log(breaks) ~ wool*tension, data = warpbreaks)
check_model(m1, check = c("linearity", "qq"), detrend=FALSE, alpha=0)
```

::: notes
standard checks we run to verify the distribution of y, the dependent variable
:::

## Example: Survival Times of Cancer Patients

Survival: $y_i \sim Exp(\mu_i)$

$$ \text{log }\mu_i =  \beta_0 + \beta_1 \text{ log } x_i  $$

Quantile residuals:

$$ r_{i} = \Phi^{-1} \left\{  1-exp(y_i/\hat{\mu_i}) \right\}$$

## GLMM Formulation: Binomial Example 

Dependent variable: $Y_{ij}$ (e.g. survival)

survival: $p_i = Y_i/N$

$$Y_{ij}|r_j \sim Binomial(N, \pi_{ij})$$

$$\eta_{ij} = \eta + \alpha_i + r_j$$
($\eta$ = mean, $\alpha_i$ = treatment, $r_j$ = grouping variable)

$$\eta_{ij} = log(\frac{\pi_{ij}} {1 - \pi_{ij}}  )$$

::: notes
Example: survival of insects after insecticide applications ($i^{th}$ treatment) across several different locations (jth random effect/hierarchal grouping). 
:::
