---
title: "Using Randomized Quantile Residuals to Assess Model Fit of Generalized Linear Mixed Models"
code: 
  echo: false
---

```{r}
#| include: false
#| echo: false
library(DHARMa); library(glmmTMB); library(performance); library(dplyr); library(lme4)
```

> Randomized quantile residual was proposed in the literature [in 1996] to circumvent the...problems in traditional residuals [for generalized linear models]. However, this approach has not gained deserved awareness and attention, partly due to the lack of extensive empirical studies to investigate its performance. 
> 
> *--Feng et al, [-@Feng2017]*

## Introduction to the Problem

In the (general) linear mixed model framework,

$$\mathbf{Y = Xb + Za + \epsilon} $$

$\mathbf{Y}$ is a vector of the values for the dependent variable, $\mathbf{X}$ and $\mathbf{Z}$ are the design matrices for fixed and random effects, respectively; $\mathbf{b}$ and $\mathbf{a}$ are the slopes for the fixed effects and random effects, respectively. The model residuals, $\epsilon_i= \hat{Y_i} - Y_i$, are expected to follow a normal distribution with a mean of zero, some standard deviation and an assumption that these terms are identically and independently distributed ("iid"). These assumptions can be relaxed under certain circumstances (e.g. hetergenous variance, repeated measures/longitudinal data), but when do hold, the residuals are uncorrleated with each other and share a common variance. 


::: {.column-margin}

LMM assumptions: 

$\epsilon_i \sim N(0, \sigma^2 \mathbf{I_n}) $$
$$a \sim N(0, \sigma_a^2) $$

$$ y_i|a_i \sim𝑁(x_i \beta, \sigma^2/r_i) $$
:::

These assumptions can be quickly assessed by plotting the residuals versus the fitted values and generating a qq-plot. 

```{r}
#| label: fig-ex1
#| fig-cap: "A residual pattern from an LMM"
#| column: margin
#| echo: false
library(lme4)
data("sleepstudy")
m1 <- lmer(Reaction ~ Days + (Days|Subject), data = sleepstudy)
plot(m1, form = resid(., type = "response") ~ fitted(.))
```

It is difficult to examine modelling assumptions in GLMMs using the same standard residual plots developed for general linear models. Model residuals in GLMMs are not expected to be normally distributed (the log-normal model being an exception) and may have no distributional assumptions at all. Furthermore, canonical relationship between mean and variance for some distributions as well as distributions for discrete outcomes can result in non-sensical plots that lack diagnostic capabilities for GLMMs. 

Alternatives to the standard residuals have been developed included Pearson, studentized and deviation residuals. 

::: {.callout-note collapse="true"}
## Types of residuals

**Raw:**
$$\epsilon_i = Y_i - \hat{Y_i} $$

**Pearson/Scaled:**

$$ \epsilon_i = \frac {Y_i - \hat{Y_i}} {sd(Y_i)}  $$

**Deviance**

$$ \epsilon_i = sign(Y_i - \hat{Y_i})\sqrt{d_i}  $$
and

$$ d_i = \frac {Y_i - \hat{Y_i}} {\sigma} $$
*(The Pearson residual for a normally-distributed variable)*
:::

These take different scaling approaches in order to decouple the relationship between the fitted values and the residual variance. Plots of residuals of all types (e.g., Pearson, deviance) from GLMMs are less ammenable to direct interpretation. In such cases, detecting deviations from model expectations such as overdisperson, underdispersion or zero-inflation is difficult, and standard goodness of fit tests are not valid. 

```{r}
#| label: fig-ex2b
#| fig-cap: "Residual plots for a binomial-distributed variable"
#| echo: false
library(lme4)
data("VerbAgg")
m2 <- glmer(r2 ~ (Anger + Gender + btype + situ)^2 + (1|id) + (1|item), 
            family = binomial, data = VerbAgg, nAGQ=0L)

plot(m2, form = resid(., type = "deviance") ~ fitted(.))
```

 Since conventional residual plots are useless for GLMMs, what tools do work? And more saliently, what modelling assumptions of GLMMs should practioners evaluate? 

## Randomized Quantile Residuals

Randomized quantile residuals are an alternative type of residuals specifically developed for generalized linear models. They were first defined by Dunn and Smyth [-@dunn1996]: and later implemented in their package "statmod" [-@statmod]. They are calculated by simulating data for each data point using the fitted GLMM for the distribution and its parameters and obtaining the quantile for each observation from the cumulative distribution function of the simulated data. Thus, quantile residuals are a measure of how well the data conform to the specified model and distribution. 

Quantile residuals are calculated following this procedure:

For $y_1,...,y_n$ responses, they follow a probability distribution with parameters $\mu$ and $\phi$. 
$$y_i \sim \mathcal{P}(\mu_i, \phi)$$

Under the cumulative distribution function for that probabality distribution, $F(y; \mu, \phi)$, the residuals, $u_i$ are calculated by sampling within an observation-specific uniform distribution and then calculating the z-score from that. 

Residuals: $r_{i,q} = \Phi^{-1}(u_i)$

Residual quantiles: $ u_i \sim Uniform(a_i, b_i]$

Where: 

$$a_i = lim_{y \rightarrow y_i} F(y; \hat{\mu_i}, \hat{\phi})$$

$$ b_i = F(y_i; \hat{\mu_i}, \hat{\phi}) $$

On a practical level, $a_i$ and $b_i$ can be described as thus: 

For a discontinous $F$: 

- $a_i = Pr(Y < y_i)$
- $b_i = Pr(Y <= y_i)$
- (a, b]

For a continuous $F$:

- $a=b$  
- $F(y_i; \mu_i, \phi)$ are uniformly distributed if the model is correct; and   

The quantile residuals, $r_i$ converge to a normal distribution when parameters are consistently estimated, and under larage samples, are iid. The Probit distribution,  $\Phi(x)$ returns a probability p, and $\Phi^{-1}(p)$ returns the z-score. The process, first described by Dunn and Smyth [-@dunn1996], was later expanded upon by Warton et all [-@warton2017] and named "probablity integral transform" residuals (abbreviated "PIT" by the authors). While Dunn and Smyth's version uses a half open interval for (a, b], Warton et al's method is closed [a, b]. 

> Randomization is used to produce continuously distributed residuals when the response is discrete or has a discrete component. This means that the quantile residuals will vary from one realization to another for a given data set and fitted model.
> *--Dunn & Smythe (-@Dunn1996)*

Dunn and Smythe originally recommend running *at least* 4 quantile samples per observation. Nowadays, the current recommendation (and default behavior of most software) is to run hundreds of simulations per observation. 

Quantile residuals can be interpreted similarly to common diagnostic residuals plots such as the q-q plot and the residual-versus-fitted values plot. Thus, randomized quantile residuals permit users to evaluate how well the data conform to model distributional assumptions and model parameter expectations. 

This method is implemented in the R packages "statmod" for `glm` objects [@statmod] and  by "DHARMa" for many linear modelling object types [@dharma]. Since DHARMa has broad compatibility with many GLMM modelling packages and is widely used by reseachers, funcationality from that package will be the focus for the rest of this chapter. 

### DHARMa (Diagnostics for HierArchical Regression Models)

DHARMa is a widely-used package for GLMM model diagnostics. It implements quantile residuals as described by Dunn and Smythe with a few alterations already aluded to: 

- the residuals are kept on the uniform distribution bounded by zero and 1 by default; they are not back-tranformed to normally-distributed Z-scores automatically, although there is functionality within the package to that. 
- Unlike the Dunn and Smyth quantile residuals, both ends of the distribution are closed: [0,1]. 
- in the default settings, 250 simulations are run for each observation in the data set

### The DHARMa Process

1. Model a process using a generalized linear model with a chosen distribution and link function. 

$$\mathbf{Y = X\beta+Za}$$
$$ \mathbf{Y|a\sim G(\mu, R)} $$

"G" is generalized term denoting any distribution with shape and scale parameters

linear predictor: $\mathbf{\eta = X\beta}$

fitted value: $E(Y) = \mu$

Link function: $\eta = g(\mu)$


2. *For each observation in the data set*, simulate new observations using fitted value as the mean and model parameters (e.g. shape, scale) for the other distributional parameters.

$$ Y_i \sim G(\hat{\mu_i}, \hat{\phi}) $$

3. Calculate the quantile for the cumulative distribution function 

$$ r_{i} = F(y_i; \hat{\mu_i}, \hat{\phi} )$$

This process is valid for Gaussian-distributed variables. DHARMa runs 250 simulations per observation by default; the documentation recommends running up to 1000 simulations per observation in the data set. 

### Expectations of the Quantile Residuals

$$ r_{ij} \sim Uniform(0,1) $$

$r_{ij} = 0$:   everything is larger  

$r_{ij} = 1$:   everything is smaller 

$r_{ij} = 0.5$:   right in the middle   


### Quantile Residuals for a Poisson-Distributed Variable

Assume we have a Poisson-distributed variable with the following conditions: 

$$ \hat{Y} = \lambda = 5, \quad Y_i = 7$$
In a data set with 500 simulated observations, we can calculate the quantile for an observed value of 7 given the mean and variance are 5. 

```{r, echo=FALSE}
#| fig-cap: "A randomized quantile residual for a single observation"
cex_axis = 1.1
cex_lab  = 1.3

par(mar = c(5.1, 5, 2.1, 2.1), mfrow = c(1, 2))

yi = 7; mu = 5
x <- rpois(500, mu)

dat = as.data.frame(table(x)) |> 
  mutate(x = as.integer(x)) |> arrange(x) |> 
  mutate(tot = cumsum(Freq),
         dens = tot/sum(Freq))

res = filter(dat, x == yi) |> pull(dens)  

hist(x, main = NULL, freq = FALSE, 
      cex.axis = cex_axis, cex.lab = cex_lab,
     xlab = expression(paste("simulated values for ", hat(Y[i]))),
     ylab = "Density", col = "#AFE4DE")
abline(v = yi + .5, lwd = 2, lty = 2)

plot(dat$x, dat$dens, col = "red2", lty = 1, 
     cex.axis = cex_axis, cex.lab = cex_lab,
     xlab = expression(paste("simulated values for ", hat(Y[i]))), 
     ylab = "Cumulative Density", type = "l", lwd = 1.5) 

lines(x = c(yi,yi), y = c(0, res), lwd = 2, lty = 2)
lines(x = c(0,yi), y = c(res, res), lwd = 2, lty = 2)
text(8, 0.4, expression(paste(Y[i])))
text(4, 0.8, expression(paste(r[i])))

rm(yi, x, dat, res, mu)
```

For a single observation, a range of quantiles residuals will result from the simulations. Below is an example of different residuals that results for the conditions of $F(y_i; \hat{\mu}) \sim Poisson(\mu)$ where $\hat{\mu} = 5$ and $y_i = 7$. 

```{r, echo=FALSE}
# generate multiple simulations to see
res_sim <- function(mu = 5, yi = 7) {
  x1 <- rpois(500, mu)
  dat1 = as.data.frame(table(x1)) |> 
    arrange(x1) |> 
    mutate(x1 = as.integer(x1)) |> arrange(x1) |> 
    mutate(tot = cumsum(Freq),
           dens = tot/sum(Freq))
  filter(dat1, x1 == yi) |> pull(dens) 
  }

#res_sim(mu = 5, yi = 7)

sims <- replicate(500, res_sim())

par(mfrow = c(1, 1))

hist(sims, main = expression(paste("Poisson Example: ", hat(Y[i]), " = ", lambda, " = 5, ", 
                                   Y[i], " = 7")),
     xlab = "Quantile Residuals (500 simulations of 500 obs each)", col = "#AFE4DE")
abline(v = mean(sims), lwd = 2, lty = 2, col = "red2")
```



### DHARMa example for a GLMM with a Poisson-distributed variable 

It always helps to see an example. Below, the DHARMa package itself is used to simulate a data set with a single random effect of "group" (10 levels), a fixed effect "Environment1", and the Poisson-distributed response variable, "observedResponse", with a scale parameter of 3 and no overdispersion. 

```{r, echo=TRUE}
library(DHARMa)

testData = createData(sampleSize = 500, 
                      fixedEffects = 1,
                      numGroups = 10,
                      randomEffectVariance = 1,
                      overdispersion = 0, 
                      family = poisson(), scale = 3,)
head(testData)
```

Distribution of the simulated dependent variable look as expected for a Poisson-distributed variable: 

```{r}
par(mfrow = c(1, 1))
hist(testData$observedResponse, col = "#AFE4DE", main = NULL, xlab = "raw data")
```

Next, the data set is analyzed correctly using a GLMM and incorrectly using a LMM: 

```{r}
#| echo: TRUE
library(lme4)

rightModel <- glmer(observedResponse ~ Environment1 + (1|group) , 
                     family = "poisson", data = testData)

wrongModel <- lmer(observedResponse ~ Environment1 + (1|group) , 
                     data = testData) 
```

Here are the standard residual plots for both deviance and Pearson residuals. Without the labels, it is difficult to ascertain the correct model. 

```{r}
#| echo: TRUE
#| fig-cap: "Standard residual plots for Poisson-distributed variable"
par(mfrow = c(1, 2))

plot(fitted(rightModel), residuals(rightModel, type = "deviance"), main = "correct family", xlab = "fitted", ylab = "deviance residuals", col = "#5E48FF")
plot(fitted(wrongModel), residuals(wrongModel, type = "pearson"), main = "incorrect family", xlab = "fitted", ylab = "Pearson residuals", col = "#5E48FF")
```

Now, let's Look at the DHARMa quantile residual output for the correctly specified model:

```{r}
#| echo: TRUE
#| fig-cap: "Randomized quantile residuals for a correctly specified model"
sr <- simulateResiduals(rightModel, plot = TRUE)
```

And also see DHARMa quantile residual output for the incorrectly specified model:


```{r}
#| echo: TRUE
#| #| fig-cap: "Randomized quantile residuals for an incorrectly specified model"
wr <- simulateResiduals(wrongModel, plot = TRUE)
```

There is a considerable amount of information here, so let's go through it one element at a time. 

### Default DHARMa Output

These plots are the results of the previously described randomized quantile residual simulations. By default, 250 simulations are conducted per observation, and simulations are conducted at the last stochastic level (i.e. across all random effects that enter the Poisson process. The simulations are saved to an object so they can re-used if necessary for model diagnostic tests and plots such as for over- or underdispersion, uniformity, zero inflation, and categorical dependence among the independent variables.

The **qq-plot** on the left is comparing observed residual "quantiles" (these are actually probabilies) with expected quantiles that were generated assuming a uniform distribution (0, 1) simulated with `stats::punif()`.   This plot can be individually called with `plotQQunif()`. It is automatically testing for uniformity, outliers and dispersion. Uniformity is evaluted with the KS (Kolmogorov-Smirnov) test, using the function `stats::ks.test()`.

The default "Dispersion test" present the results of a DHARMa bootstrap approach, comparing the variance of the observed raw residuals versus the variance of the simulated residuals (i.e. simulated - predicted values), where both are scaled to the mean simulated variance. Values close indicate conformation to expectations; values greater than one provide evidence of overdispersion and values less than one indicate underdispersion. This is the process for a fully conditional model. An unconditional model uses the squared Pearson residuals to calculate and compare approximate deviances. This can be computationally intensive and requires that the modelling package furnish Pearson residuals. In both cases, the p-value is the empirical proportion a simulated values is greater, less than or both compared to the observed value. Another approach is enabled using Pearson residuals described by Ben Bolker as a "crude" and  "approximate" estimation in [GLMM FAQ's](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#overdispersion). This uses the squared sum of the Pearson residuals and error degrees of freedom in the Chi-square density function to estimate the proability of over- or under-dispersion.  All dispersion tests implemented in DHARMa are by default two-sided, evaluating both overdispersion and underdispersion. This can be adjusted in the function arguments if only one approach is needed. 

the **Outlier Test** is a bit tricky to interpret. DHARMa defines "outliers" as any observed value that is zero (all simulated data are larger) or 1 (all simulated data are smaller);^[This is one reason why transforming these with the quantile function can be a misleading since probabilities of 0 and 1 end up as negative infinity and infinity, respectively. Amusingly, the outlier test is a 2-sided test. "Are there enough outliers?" is not something researchers ask frequently.] thus, no information on magnitude is given. The outlier test runs a binomial test where the expected number of outliers is the inverse of the number of simulations plus 1 ($1/(nSim + 1)$). The DHARMa manual recommends that users view data labelled as "outliers" by DHARMa with caution as there is no way to determine to what extent they are truly outlying values. There is also a bootstrap procedure that empirically calculates the number of expected outliers versus how many were observed in the data. 

The plot on the right is contrasting the quantile residuals versus the rank-transformed model predictions. DHARMa conducts quantile tests at the 0.25, 0.5 and 0.75 quantiles by fitting splines (using the function `qgam::qgam()`) and testing for deviations from the expectation of a flat line. It does a Benjamini-Hochberg adjustment for multiple testing. This plot can be called with the function `DHARMa::plotResiduals()`, that has additional arguments for using absolute deviations (for testing heteroscedasticity), testing versus other specific factors in the model, and other arguments.  [[add content on plot interpretation]]

### Additional DHARMa Functionality to Consider

**1. Testing for specific factor**
Residuals can be simulated for individuals predictors to evaluate the impact of specific factor levels on dispersion using the `form = ` argument in `plotResiduals()` or the function `testCategorical()`). In such cases, boxplots are produced, betweeen group variance is compared with a Levene's test, and within-group variance is evaluted with individual KS tests (simulating a uniform distribution for each factor level). 

**1. Test for zero-inflation**

A handy test evaluating how the number of zero's occurring measure up to the number of zero's expected for a given distribution and set of parameters. Like the other tets, the number of zero's under the simulations are compared to the observed zeros and an empirical probability of having more or less zero's than expected is reported. 

**1. Tests for spatial and temporal autocorrelation**

Note that for any kind of autocorrelation, conditional simulations are recommendfed otherwise the plots or tests may indicate problems when there are not any. 
[[need to finish content]] 

**1. Simulate Likelihood Ratio Ttest**

[[need content]]

*Please note that this is not a comprehensive list of all functionality enabled in DHARMa.*

### Conditional and Marginal Models

The default choice of DHARMa is to simulate all stochastic levels, so the resulting residuals can serve an omnibus test. The documentation for the package indicates that if the model is correctly specified, the simulated quantile residuals are diagnostic regardless of how many hierarchal levels simulated.

Given the tricky situation where low level random effects enter the model as Gaussian variables that are subsumed into the GLMM process, there may be interest in simulating lower level processes. The extent to which these simulations can be conducted depends on the R package being used. Conditional models where the predictions are conditioned on the random effects only are fully supported for **lme4** and partially supported for **glmmTMB** (see [GitHub issue](https://github.com/glmmTMB/glmmTMB/issues/888) for technical details). The DHARMa documentation warns that unconditional simulations can result in greater variability of the residuals and lower power to detect deviations from model expectations in higher-level processes. In particular, dispersion tests can be less sensitive using unconditional rather than conditional simulations. 

Below are examples of conditional and unconditional residual simulations. 

[[finish content]]

```{r, echo=TRUE,eval=FALSE}
m2 <- glmer(r2 ~ (Anger + Gender + btype + situ)^2 + (1|id) + (1|item), 
            family = binomial, data = VerbAgg, nAGQ=0L)
```

```{r}
r2 <- simulateResiduals(m2, plot=TRUE, n = 50)
```

## Limitations to Randomize Quantile Residuals

There is a relative paucity of peer-reviewed literature. [[finish content]]


## Conclusion and Best Practices

Assessing if a generalized linear mixed model (GLMM) is appropriately specified for a given data set and data generating process is challenging since often GLMMs have no defined distribution for model residuals. Randomized quantile residuals are a useful model diagnostic tool for those situations. This is implemented in the R package DHARMa and be manually written for SAS. 

- **Randomized quantile residuals are a great tool.** These are helpful for diagnosing model misspecification and do work much better than conventional residuals for identifying problems. 

- **Interpret DHARMa results as suggetions, not rules.** Although the hypothesis tests conducted by DHARMa provide an easy decision making tool, recall that no hypothesis provides an unambiguous answer. Outlier test results in particular should be treated with caution and not strictly diagnostic for individual observations.

- **Functionality** in DHARMa depends on the simulation functions build into GLMM package that is being supported and what options are available through that package. Supported packages and function include: `lm()`, `glm()`, **lme4**, **glmmTMB**, **GLMMadaptive**, **mgcv**, **spaMM**, **mgcv**, **gamm4**, **brms**, among other packages. 

- **Pay attention to other tests.** When appropriate, make sure to evaluate candidate models for heteroscedasticty due to individual factors in the model, proper specification of a zero-inflation terms and autocorrelation (spatial or temporal). These are model specific and are not produced by the default DHARMa output, so the user must be aware and motivated. 

- **Conditional simulations** Commonly only the last stochastic level (e.g. Poisson) is simulated, conditional on the fitted random effects, serving as an omnibus test for model residuals. Unconditional simulations are dependent on the GLMM package itself and are implemented to varying extents in each package. These may be useful in certain scenarios, but overall lack diagnostic power. 

- **Be aware of limitations.** It should be noted that at this time, there is a lack of information on their performance and reliability across a broad range of conditions. The sensitivity of these residuals to modest model mis-specifications is unknown. 

-  **Categorical models** randomized quantile residuals do not work with ordinal or multinomial models. 

- **Read the manual.** The documentation for DHARMa is extensive and detailed; reading this will enhance user's understanding of how to best use DHARMa and interpret its output. 
 

----------------------------------------------------------------

## Cut Material

*If you think anything below is worth keeping, let me know.*


** This stuff is from the Dunn & Smythe paper:**

::: {.callout-note collapse="true"}
## Example: survival times of patients 
*(a continuous variable)*

Survival: $y_i \sim Exp(\mu_i)$

$$ \text{log }\mu_i =  \beta_0 + \beta_1 \text{ log } x_i  $$

Quantile residuals:

$$ r_{i} = \Phi^{-1} \left\{  1-exp(y_i/\hat{\mu_i}) \right\}$$

For Discontinuous CDF, $F$

$$r_i = \Phi^{-1}(u_i)$$
$$ u_i \sim Uniform(a_i, b_j]$$
$$a_i = lim_{y \uparrow y_i} F(y; \hat{\mu_i}, \hat{\phi})$$

$$ b_i = F(y_i; \hat{\mu_i}, \hat{\phi})  $$

:::

::: {.callout-note collapse="true"}
## Example: binomial trait

$$y_i \sim binomial(n, p_i) $$

$$ \text{logit} (p_i) = \beta_0 + \beta_1x_i $$
$$r_i = \Phi^{-1}(u_i)$$

$$ u_i \sim Uniform(a_i, b_j] $$

$$a_i = lim_{y \uparrow y_i} F(y; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k-1 \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$


$$b_i = F(y_i; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$
:::


