---
title: "Using Randomized Quantile Residuals to Assess Model Fit of Generalized Linear Mixed Models"
---

```{r}
#| include: false
library(DHARMa); library(glmmTMB); library(performance); library(dplyr); library(lme4)
```

> Randomized quantile residual was proposed in the literature [in 1996] to circumvent the...problems in traditional residuals [for generalized linear models]. However, this approach has not gained deserved awareness and attention, partly due to the lack of extensive empirical studies to investigate its performance. 
> 
> *--Feng et al, [-@Feng2017]*


Assessing if a generalized linear mixed model (GLMM) is appropriately specified for a given data set and data generating process is challenging since often GLMMs have no defined distribution for model residuals. In a (general) linear mixed modelling context under the following conditions:

$$\mathbf{Y = Xb + Za + \epsilon} $$
$$\epsilon_i \sim N(0, \sigma^2 \mathbf{I_n}) $$
$$a \sim N(0, \sigma_a^2) $$
$$ y_i|z_i \sim𝑁(x_i \beta, \sigma^2/r_i) $$

The residuals, $\epsilon_i= \hat{Y_i} - Y_i$ are expected to follow a normal distribution with a mean of zero, some standard deviation and an assumption that these terms are identically and independently distributed. These assumptions can be relaxed under certain circumstances (e.g. hetergenous variance, repeated measures), but when do hold, the residuals are uncorrleated with each other and share a common variance. This can be quickly assessed by plotting the residuals versus the fitted values. 

```{r}
data(warpbreaks)
m1 <- lm(log(breaks) ~ wool*tension, data = warpbreaks)
check_model(m1, check = c("linearity", "qq"), detrend=FALSE, alpha=0)
```

Raw residuals from GLMMs can be nonsensical, and overall lack diagnostic capabilities for GLMMs. 

```{r, echo=TRUE}
data("VerbAgg")
m2 <- glmer(r2 ~ (Anger + Gender + btype + situ)^2 + (1|id) + (1|item), 
            family = binomial, data = VerbAgg, nAGQ=0L)
plot(m2)
```

Detecting deviations from model expectations (e.g. dispersion, zero-inflation) is difficult given that some non-normal distributions have mathematically defined mean-variance relationships, and they may have discrete outcomes that are less amenable to direct interpretation via plotting. The visual patterns are difficult to interpret, over and under-dispersion are difficult to assess, and standard goodness of fit tests are not valid. 

Scaled residuals such as Pearson or studentized residuals can alleviate these challenges to an extent, but they also can be difficult to correctly interpret for over- and underdispersion. 

::: {.callout-note collapse="true"}
## Residuals 

**Raw:**
$$\epsilon_i = Y_i - \hat{Y_i} $$

**(internally) Studentized:**

$$\epsilon_i = \frac {Y_i - \hat{Y_i}} {\hat{s}}$$

**Pearson/Scaled:**

$$ \epsilon_i = \frac {Y_i - \hat{Y_i}} {sd(Y_i)}  $$
:::

## Randomized Quantile Residuals

Randomized quantile residuals are an alternative, simulating data for each data point using the fitted GLMM for the distribution and its parameters and obtaining the quantile for each observation from the cumulative distribution function of the simulated data. Thus, quantile residuals are a measure of how well the data conform to the specified model and distribution. 

Quantile residuals were first defined by Dunn and Smyth [-@dunn1996]: and later implemented in their package "statmod" [-@statmod]. 

For $y_1,...,y_n$ responses:  

$$y_i \sim \mathcal{P}(\mu_i, \phi)$$

CDF: $$F(y; \mu, \phi)$$ 

For a continuous $F$, $F(y_i; \mu_i \phi)$ are uniformly distributed, and the quantile residuals are:

$$ r_{i} = \Phi^{-1} \left\{F(y_i; \hat{\mu_i}, \hat{\phi} )\right\}$$

The quantile residuals, $r_i$ converge to a normal distribution when parameters are consistently estimated. The Probit distribution,  \Phi(x) returns a probability p, and \Phi^{-1}(p) returns the Z-score. 

#### Example: survival times of patients (continuous)

Survival: $y_i \sim Exp(\mu_i)$

$$ \text{log }\mu_i =  \beta_0 + \beta_1 \text{ log } x_i  $$

Quantile residuals:

$$ r_{i} = \Phi^{-1} \left\{  1-exp(y_i/\hat{\mu_i}) \right\}$$

### Discontinuous CDF, *F*

$$r_i = \Phi^{-1}(u_i)$$
$$ u_i \sim Uniform(a_i, b_j]$$
$$a_i = lim_{y \uparrow y_i} F(y; \hat{\mu_i}, \hat{\phi})$$

$$ b_i = F(y_i; \hat{\mu_i}, \hat{\phi})  $$



#### Example: binomial trait

$$y_i \sim binomial(n, p_i) $$

$$ \text{logit} (p_i) = \beta_0 + \beta_1x_i $$
$$r_i = \Phi^{-1}(u_i)$$

$$ u_i \sim Uniform(a_i, b_j] $$

$$a_i = lim_{y \uparrow y_i} F(y; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k-1 \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$


$$b_i = F(y_i; n, \hat{p_i}) =\displaystyle\sum_{i=0}^{\lfloor k \rfloor} \begin{pmatrix}
  n \\
  i
\end{pmatrix} p_i(1-p)^{n-i}$$


> Randomization is used to produce continuously distributed residuals when the response is discrete or has a discrete component. This means that the quantile residuals will vary from one realization to another for a given data set and fitted model.
> *--Dunn & Smythe (-@Dunn1996)*

They originally recommend running *at least* 4 quantile samples per observation. 

Quantile residuals are expected to follow a uniform distribution with a half-open interval (0,1], and they can be interpreted similarly to common diagnostic residuals plots such as the q-q plot and the residual-versus-fitted values plot.  This method is implemented in the R packages "statmod" for `glm` objects [@statmod], and an adapted version is implemented by "DHARMa" for many linear modelling object types [@dharma]. 

## DHARMa (Diagnostics for HierArchical Regression Models)

DHARMa is a widely-used package for GLMM model diagnostics. It implements quantile residuals as described by Dunn and Smythe with a few alterations: 

- the residuals are kept on the uniform distribution bounded by zero and 1; they are not back-tranformed to normally-distributed Z-scores. Unlike the Dunn and Smyth quantile residuals, both ends of the distribution are closed: [0,1]. 
- in the default settings, 250 simulations are run for each observation in the data set
- for discrete outcomes, a 


### The DHARMa Process

1. Model something using a generalized linear model with a chosen distribution and link function. 


$$\mathbf{Y = X\beta+Za}$$
$$ \mathbf{Y|a\sim G(\mu, R)} $$
linear predictor: $\mathbf{\eta = X\beta}$

fitted value: $E(Y) = \mu$

Link function: $\eta = g(\mu)$


"G" is generalized term denoting any distribution with shape and scale parameters


2. *For each observation in the data set*, simulate new observations using fitted value as the mean and model parameters (e.g. shape, scale) for the other distributional parameters.

$$ Y_i \sim G(\hat{\mu_i}, \hat{\phi}) $$

3. Calculate the quantile for the cumulative distribution function 

$$ r_{i} = F(y_i; \hat{\mu_i}, \hat{\phi} )$$

This process is valid for Gaussian-distributed variables. DHARMa runs 250 simulations per observation by default; the documentation recommends running up to 1000 simulations per observation in the data set. 

### Expectations of the Quantile Residuals

$$ r_{ij} \sim Uniform(0,1) $$


$r_{ij} = 0$:   everything is larger  

$r_{ij} = 1$:   everything is smaller 

$r_{ij} = 0.5$:   right in the middle   


### Distribution of Quantile Residuals for a Poisson-Distributed Vairabale

Say we have a Poisson-distributed variable with the following conditions: 

$$ \hat{Y_i} = \lambda = 5, \quad Y_i = 7$$
In a data set with 500 simulated observations, we can calculate the quantile for an observed value of "7" given the mean and variance are "5". 

```{r, echo=FALSE}
cex_axis = 1.1
cex_lab  = 1.3

par(mar = c(5.1, 5, 2.1, 2.1), mfrow = c(1, 2))

yi = 7; mu = 5
x <- rpois(500, mu)

dat = as.data.frame(table(x)) |> 
  mutate(x = as.integer(x)) |> arrange(x) |> 
  mutate(tot = cumsum(Freq),
         dens = tot/sum(Freq))

res = filter(dat, x == yi) |> pull(dens)  

hist(x, main = NULL, freq = FALSE, 
      cex.axis = cex_axis, cex.lab = cex_lab,
     xlab = expression(paste("simulated values for ", hat(Y[i]))),
     ylab = "Density", col = "#AFE4DE")
abline(v = yi + .5, lwd = 2, lty = 2)

plot(dat$x, dat$dens, col = "red2", lty = 1, 
     cex.axis = cex_axis, cex.lab = cex_lab,
     xlab = expression(paste("simulated values for ", hat(Y[i]))), 
     ylab = "Cumulative Density", type = "l", lwd = 1.5) 

lines(x = c(yi,yi), y = c(0, res), lwd = 2, lty = 2)
lines(x = c(0,yi), y = c(res, res), lwd = 2, lty = 2)
text(8, 0.4, expression(paste(Y[i])))
text(4, 0.8, expression(paste(r[i])))

rm(yi, x, dat, res, mu)
```

For a single observation, there is an observable range of quantiles that can result: 

```{r, echo=FALSE}
# generate multiple simulations to see
res_sim <- function(mu = 5, yi = 7) {
  x1 <- rpois(500, mu)
  dat1 = as.data.frame(table(x1)) |> 
    arrange(x1) |> 
    mutate(x1 = as.integer(x1)) |> arrange(x1) |> 
    mutate(tot = cumsum(Freq),
           dens = tot/sum(Freq))
  filter(dat1, x1 == yi) |> pull(dens) 
  }

#res_sim(mu = 5, yi = 7)

sims <- replicate(500, res_sim())

par(mfrow = c(1, 1))

hist(sims, main = expression(paste("Poisson Example: ", hat(Y[i]), " = ", lambda, " = 5, ", 
                                   Y[i], " = 7")),
     xlab = "Sample Quantiles (500 simulations of 500 obs each)", col = "#AFE4DE")
abline(v = mean(sims), lwd = 2, lty = 2, col = "red2")
```

### DHARMa example for a GLMM with a Poisson-distributed variable 

The package can create a data set with a random effect ("group", 10 levels), a covariate ("Environment1") and a Poisson-distributed response variable ("observedResponse"). The default distribution is Poisson with a scale parameter of 1. 

```{r, echo=TRUE}
testData = createData(sampleSize = 500)
head(testData)
```

Distribution of the simulated data: 

```{r}
par(mfrow = c(1, 1))
hist(testData$observedResponse, col = "#AFE4DE", main = NULL, xlab = "raw data")
```

Analyze correctly using a GLMM and incorrectly using a LMM:

```{r, echo=TRUE}
rightModel <- glmer(observedResponse ~ Environment1 + (1|group) , 
                     family = "poisson", data = testData)

wrongModel <- lmer(observedResponse ~ Environment1 + (1|group) , 
                     data = testData)
```


Examine the 'Standard' Residuals

```{r}
par(mfrow = c(1, 2))

plot(fitted(rightModel), residuals(rightModel, type = "deviance"), main = "correct family", xlab = "fitted", ylab = "deviance residuals", col = "#5E48FF")
plot(fitted(wrongModel), residuals(wrongModel, type = "pearson"), main = "incorrect family", xlab = "fitted", ylab = "Pearson residuals", col = "#5E48FF")
```

Look at the quantile residuals for the correctly specified model:

```{r}
sr <- simulateResiduals(rightModel, plot = TRUE)
```

Many hypothesis tests can be conducted using quantile residuals including tests for over- or underdispersion, uniformity, zero inflation, and categorical dependence among the independent variables.

Compare the quantile residuals for the incorrectly specified model:


```{r}
wr <- simulateResiduals(wrongModel, plot = TRUE)
```


### DHARMa Output

The qq-plot on the left is comparing observed residual "quantiles" (these are actually probabilies) with expected quantiles that were generated assuming a uniform distribution (0, 1) simulated with `stats::punif()`. The "KS Test" is a Kolmogorov-Smirnov test, using the function `stats::ks.test()`.   

The "Dispersion test" is ....

the "Outlier Test" is a bit tricky to interpret. DHARMa defines "outliers" as any observed value that is zero (all simulated data are larger) or 1 (all simulated data are smaller);^[This is one reason why transforming these with the quantile function can be a misleading since probabilities of 0 and 1 end up as negative infinity and infinity, respectively. Amusingly, the outlier test is a 2-sided test. "Are there enough outliers?" is not something people ask frequently.] thus, no information on magnitude is given. The outlier test runs a binomial test where the expected number of outliers is the inverse of the number of simulations plus 1 (1/(nSim + 1)). The DHARMa recommends that users view data labelled as "outliers" by DHARMa with caution as there is no way to determine to what extent they are truly outlying values. There is also a bootstrap procedure that empirically calculates the number of expected outliers versus how many were observed in the data. 

The plot on the right is contrasting the quantile residuals versus the 
- quantile tests: fits splines (`qgam::qgam()`) at 0.25, 0.5 and 0.75 quantiles and looks for deviations from the expectation of a flat line, does a BH adjustment




### Additional DHARMa Tests
bn 

### Notes on DHARMa Implementation 

Functionality in DHARMa depends on the simulation functions build into GLMM package that is being supported and what options are available through that package. Commonly, only the last stochastic level (e.g. Poisson) is simulated, conditional on the fitted random effects, basically an omnibus test. 
- Residuals can be simulated for individuals predictors and tests can be conducted for individual factors
- for any kind of autocorrelation, conditional simulations are a must so that plots don't indicate patterns or the tests fail due to autocorrelation
- Supported packages/function: `lm()`, `glm()`, **lme4**, **glmmTMB**, **GLMMadaptive**, **mgcv**, **spaMM**, **mgcv**, **gamm4**, **brms**, and more



> If the model is correctly specified, the simulated residuals should be flat regardless how many hierarchical levels we re-simulate. The most thorough procedure would therefore be to test all possible options. If testing only one option, I would recommend to re-simulate all levels, because this essentially tests the model structure as a whole. This is the default setting in the DHARMa package. A potential drawback is that re-simulating the lower-level random effects creates more variability, which may reduce power for detecting problems in the upper-level stochastic processes. In particular dispersion tests may produce different results when switching from conditional to unconditional simulations, and often the conditional simulation is more sensitive.
> *--F. Hargtig, DHARMA Manual* [-@dharma]

### Conditional and Marginal Models

Conidtional models where the predictions are conditioned on the random Only fully supported for **lme4**, partially supported for **glmmTMB** (see [GitHub issue](https://github.com/glmmTMB/glmmTMB/issues/888) for more details)


> A potential drawback is that re-simulating the lower-level random effects creates more variability, which may reduce power for detecting problems in the upper-level stochastic processes. In particular dispersion tests may produce different results when switching from conditional to unconditional simulations, and often the conditional simulation is more sensitive.
> *--F. Hartig, DHARMA Manual [@dharma]*


```{r, echo=TRUE,eval=FALSE}
m2 <- glmer(r2 ~ (Anger + Gender + btype + situ)^2 + (1|id) + (1|item), 
            family = binomial, data = VerbAgg, nAGQ=0L)
```

```{r}
r2 <- simulateResiduals(m2, plot=TRUE, n = 50)
```


### Other DHARMa Functionality 

* Tests for spatial and temporal autocorrelation 
* Test for zero-inflation
* Over/Under dispersion tests


## Best Practices

Randomized quantile residuals are a useful model diagnostic tool. The package documentation is extensive and detailed; reading this will enhance user's understanding of how to best use DHARMa and interpret its output.Although the hypothesis tests conducted by DHARMa provide an easy decision making tool, recall that no hypothesis provides an unambiguous answer. Outlier test results in particular should be treated with caution and not strictly diagnostic for individual observations.  It should be noted that at this time, there is a lack of information on their performance and reliability across a broad range of conditions. The sensitivity of these residuals to modest model mis-specifications is unknown. 


